Ownership types in theory and practice (in Rust)

### 2022-11-30 paper.tex
Worked on stack-based variable ownership transfer

% TODO: source
In addition to providing the memory safety that ownership types provide, ownership transfers can also reduce redundant copy operations.
Since heap-based values could be arbitrarily large at runtime, copying the whole value instead of transferring ownership could be very inefficient.

@@

% TODO: wording of value and types used interchangedly here, maybe incorrect
% TODO: maybe add defintion of ownership transfer if not obvious
\subsubsection{Transferring ownership}

@@

There are several different scenarios in which ownership is transferred.

In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
\begin{lstlisting}
let a = HeapBasedType::default();
let b = a;
\end{lstlisting}
In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
After the move, \emph{a} is no longer valid and \emph{b} is the new owner of the value.

@@

% TODO: avoid and at beginning of sentence
% And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
% Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.

@@

% TODO: bad sentence
% The following piece of code shows a \emph{String} value being \emph{moved} from one variable into another.
The heap-allocated \emph{String} type in Rust provides a good example to demonstrate the transfer of ownership of values.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
% TODO: go into detail on how strings/heap types are stored on heap.
The \emph{String} value stored in \emph{x} could be arbitrarily long at runtime.
Thus, copying the whole string into \emph{y} could be very inefficient.
% TODO: avoid and at beginning of sentence
And, since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
After the move, \emph{x} is no longer valid.
Because Rust prevents you from using the invalid reference to \emph{x} in the third line, the program will not compile.

@@

\subsubsection{Stack-based values}

First, one must differentiate between values that are stored on the stack and those stored on the heap.
% TODO: leave out immutable, adds no meaning to sentence
In Rust, primitive data types (e.g. signed 32 bit integers \emph{i32}) are immutable by default and stored on the stack, since their sizes are small and known at compile time.

% TODO: a good example of what?

% TODO: footnote that main class etc. have been omitted for sake of simplicity
% However, a similar code snippet that replaces the string literal with the \emph{String} type behaves differently.
% The following piece of code demonstrates how primitive values are \emph{moved} into other variables in Rust.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
% TODO: explain what Rust traits are
Internally, these primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.

>>>>

@@

% TODO: wording of value and types used interchangedly here, maybe incorrect
% TODO: maybe add defintion of ownership transfer if not obvious
% TODO: go into detail on how heap types are stored on heap. some constant fields are actually stored on stack (string: pointer, size, capacity)
\subsubsection{Transferring ownership}

@@

% TODO: source
In addition to the memory safety that ownership types provide, ownership transfers can also reduce redundant copy operations.
Since heap-based values could be arbitrarily large at runtime, copying the whole value instead of transferring ownership could be very inefficient.
There are several different scenarios in which ownership is transferred.

In the simplest scenario, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to \emph{a}.
% TODO: footnote that main class etc. have been omitted for sake of simplicity
\begin{lstlisting}
let a = HeapBasedType::default();
let b = a;
\end{lstlisting}
In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
After the move, \emph{a} is no longer valid and \emph{b} is the new owner of the value.

@@

The heap-allocated \emph{String} type in Rust provides a good example that demonstrates ownership transfer of heap-based values.
\begin{lstlisting}
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
% Since the \emph{String} value can only have one owner, it is \emph{moved} into \emph{y}, which will become its new owner.
After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
Therefore, Rust prevents you from accessing the invalid \emph{x} in the third line, and the program will not compile.

\paragraph{Stack-based values}

In contrast to heap-based values, the size of stack-based values is known at compile time and usually small.
Therefore, stack-based values in Rust are copied instead of moved.
% TODO: reformulate
Internally, primitive types are annotated with the \emph{Copy} trait, which specifies that variables are trivially copied instead of moved.

% The following example illustrates that the ownership of stack-based values is not transferred.
% \begin{lstlisting}
% let x = 5;
% let y = x;
% y = y + 1;
% assert_eq!(x, 5)
% \end{lstlisting}

The following piece of code demonstrates how primitive stack-based values are \emph{moved} into other variables in Rust.
\begin{lstlisting}
let x: &str = "hello world";
let y: &str = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
Since \emph{x} has a known, fixed size, its value is simply copied into \emph{y}.
Therefore, the program compiles and, as expected, "x: hello world, y: hello world" is printed to stdout.
% TODO: explain what Rust traits are

@@

\subsection{The Clone trait}

@@
